A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-08 latex/main.tex

\subsection*{Quantification of evidence for recombination}

The LS HMM at the heart of sc2ts provides a simple and powerful means 
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution for the 
putative recombinant sequence with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between
the two solutions---the number of mutations
``averted'' by recombination---is a measure of
the evidence in favor of recombination.
Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
for each of the 21 recombination events associated with Pango X lineages,
with support ranging from very strong in the case of XC
(the most parsimonious non-recombinant placement would
require 37 additional mutations) to marginal for XG and XW,
which have the minimal possible number of averted mutations.

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
\caption{
Evidence for recombination in the ARG.
All plots show QC-passing sc2ts recombination nodes,
colored according to their classification by rebar
(recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations
when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table~\ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al.\cite{Jackson2021} are prefixed with a `J', e.g.
Jackson group B (JB) and Jackson singleton recombinant 1 (J1).
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted
by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each QC-passing recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
leading to XA, XW, XBB, and an
undesignated recombination event under XBB, which we here refer to as  XBB.x.}
\label{fig:averted_mutations}
\end{figure}

Figure~\ref{fig:averted_mutations}A plots the number of mutations required
in the recombinant vs non-recombinant HMM solutions (upper scatter plot)
and the resulting distribution of the number of 
averted mutations (lower histogram)
over all 354 recombination events that pass QC. 
Of these, only 24 
(21 associated with Pango X lineages in Table~\ref{tab:pango_x_lineages},
and 3 additional from Jackon et al in Table~\ref{tab:jackson}) have 
% Slightly risky, this, as we could get asked to incorporate 
% novel recombs found by RIPPLEs? This could be done if we could get 
% Viridian sample IDs for their list.
been thoroughly characterised. 
The figure also shows the classification status for each recombinant
according to rebar, a method that classifies sequences
as recombinant or not based on groups of mutations associated with
Pango lineages (STAR Methods). 
Of the 38 recombinants that are strongly supported (at least 10 averted
mutations), 29 are novel. The known recombinants are XA, XC, XS, XBR, 
and 5 from Jackson et al. All but 2 of the 38 are classified as
recombinant by rebar. The majority (20) of these novel strongly supported
recombinants are singletons, with 9 having further descendants in the ARG.
Of the remaining 316 recombinants with less than 10 averted mutations,
only 151 are classified as recombinant by rebar. 
While some are likely to 
be false positives, it is also interesting to note that rebar validation
rate is much lower for recombinants with closely related 
parents (Figure~\ref{fig:averted_mutations}B,C) suggesting that 
sc2ts may have a higher sensitivity.

Recombination is rare in SARS-CoV-2, and ultimately judgements about
plausibility require expert analysis. 
To aid such interpretation and help identify quality control
issues we developed a visualisation 
based on SnpIt~\cite{OToole2024} and 
the recombinant view from RIVET~\cite{Smith2023tracking},
showing the informative genomic positions between 
a recombinant and its parent sequences, and highlights 
potentially problematic loci.
Each copying pattern shows the positions
where the allelic state in a recombinant
(middle row) matches that of the left (upper row, coloured green) 
or right parent (lower row, coloured blue),
or where the recombination event requires a de-novo mutation
(gold, with mutational change below).
Figure~\ref{fig:averted_mutations}D shows some example copying patterns,
illustrating this spectrum of plausibility.
XA is an example of a highly plausible recombinant, shown by the 
large number of supporting loci on the left and right. As we have 
only one de-novo mutation, we can be sure that the parents are closely
observed [Some details about XA here, how well the parents were sampled etc].
XBB is also strongly supported by many loci on the left and right, 
but the presence of many de-novo mutations
observed in the copying patterns suggests the existence of
many genetically distinct, unsampled ancestors 
[FIXME: in this case, probably BJ.1, see issue 336],
and some caution should be exercised in interpreting them. 
See Figures~\ref{fig:averted_mutations}D and
\ref{fig:pango_x_copying_patterns} for copying patterns of the 
Pango X lineages, and 
Figure~\ref{fig:copying_pattern_examples} for copying patterns
highlighting specific types of QC problem (and more details on the 
visualisation).

>>>>

\subsection*{Quantification of evidence for recombination}
The LS HMM at the heart of sc2ts provides a simple and powerful means 
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution for the 
putative recombinant sequence with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between
the two solutions---the number of mutations
``averted'' by recombination---is a measure of
the evidence in favor of recombination.
Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
for each of the 21 recombination events associated with Pango X lineages,
with support ranging from very strong in the case of XC
(the most parsimonious non-recombinant placement would
require 37 additional mutations) to marginal for XG and XW,
which have the minimal possible number of averted mutations.

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
\caption{
Properties of recombination events. A,B,C 
show summaries of the 354 sc2ts recombination nodes
passing quality control filters
colored by CovRecomb classification (recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations
when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table~\ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al.\cite{Jackson2021} are prefixed with a `J', e.g.
Jackson group B (JB) and Jackson singleton recombinant 1 (J1).
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted
by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each QC-passing recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
for XA, XW, XBB, and an
undesignated recombination event under XBB, which we refer to as XBB.x.}
\label{fig:recombinant_evidence}
\end{figure}

Figure~\ref{fig:recombinant_evidence}A plots the number of mutations required
in the recombinant vs non-recombinant HMM solutions (upper scatter plot)
and the resulting distribution of the number of 
averted mutations (lower histogram)
over all 354 recombination events that pass QC. 
Of these, only 24 
(21 associated with Pango X lineages in Table~\ref{tab:pango_x_lineages},
and 3 additional from Jackon et al in Table~\ref{tab:jackson}) have 
% Slightly risky, this, as we could get asked to incorporate 
% novel recombs found by RIPPLEs? This could be done if we could get 
% Viridian sample IDs for their list.
been thoroughly characterised. 
The figure also shows the classification status for each recombinant
according to rebar, a method that classifies sequences
as recombinant or not based on groups of mutations associated with
Pango lineages (STAR Methods). 
Of the 38 recombinants that are strongly supported (at least 10 averted
mutations), 29 are novel. The known recombinants are XA, XC, XS, XBR, 
and 5 from Jackson et al. All but 2 of the 38 are classified as
recombinant by rebar. The majority (20) of these novel strongly supported
recombinants are singletons, with 9 having further descendants in the ARG.
Of the remaining 316 recombinants with less than 10 averted mutations,
only 151 are classified as recombinant by rebar. 
While some are likely to 
be false positives, it is also interesting to note that rebar validation
rate is much lower for recombinants with closely related 
parents (Figure~\ref{fig:recombinant_evidence}B,C) suggesting that 
sc2ts may have a higher sensitivity.

See Figure~\ref{fig:recombinant_evidence_rebar} for rebar,
Figure~\ref{fig:recombinant_evidence_ripples_p4} for RIPPLES p4
and Figure~\ref{fig:recombinant_evidence_ripples_p4} for RIPPLES p3.

Recombination is rare in SARS-CoV-2, and ultimately judgements about
plausibility require expert analysis. 
To aid such interpretation and help identify quality control
issues we developed a visualisation 
based on SnpIt~\cite{OToole2024} and 
the recombinant view from RIVET~\cite{Smith2023tracking},
showing the informative genomic positions between 
a recombinant and its parent sequences, and highlights 
potentially problematic loci.
Each copying pattern shows the positions
where the allelic state in a recombinant
(middle row) matches that of the left (upper row, coloured green) 
or right parent (lower row, coloured blue),
or where the recombination event requires a de-novo mutation
(gold, with mutational change below).
Figure~\ref{fig:recombinant_evidence}D shows some example copying patterns,
illustrating this spectrum of plausibility.
XA is an example of a highly plausible recombinant, shown by the 
large number of supporting loci on the left and right. As we have 
only one de-novo mutation, we can be sure that the parents are closely
observed [Some details about XA here, how well the parents were sampled etc].
XBB is also strongly supported by many loci on the left and right, 
but the presence of many de-novo mutations
observed in the copying patterns suggests the existence of
many genetically distinct, unsampled ancestors 
[FIXME: in this case, probably BJ.1, see issue 336],
and some caution should be exercised in interpreting them. 
See Figures~\ref{fig:recombinant_evidence}D and
\ref{fig:pango_x_copying_patterns} for copying patterns of the 
Pango X lineages, and 
Figure~\ref{fig:copying_pattern_examples} for copying patterns
highlighting specific types of QC problem (and more details on the 
visualisation).
